SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-03-02 paper.tex
put more about jacobian / model size in intro

Single-Instruction, Multiple-Data (SIMD) and the related Single-Instruction, Multiple-Thread (SIMT) processing are two important vector-processing paradigms used increasingly in scientific computing.
Traditional multi-core parallelism is often used to increase central processing unit (CPU) performance, however SIMD processors---as well as SIMT processors, e.g., in the form of graphics processing units (GPUs)---have gained recognition due to their increased floating operation throughput.
The parallel programming standard OpenCL~\cite{stone2010opencl} has further enabled adoption of vector-processing based codes in scientific computing by providing a common application program interface (API) for execution on heterogeneous systems, e.g., the CPU, GPU, or Intel's Many Integrated Core architecture (MIC), etc.
In this discussion, we will largely use OpenCL terminology to describe these processing paradigms, as it provides a convenient way to classify otherwise disparate processor type (e.g., CPUs and GPUs), however the concepts discussed are broadly applicable to SIMD\slash SIMT processing and not tied to OpenCL specifically.

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMD.pdf}
      \caption{Schematic of SIMD processing.  A single compute unit (e.g., a CPU core) contains a vector unit with $N_v$ processing elements, together called a vector-lane.  The vector unit executes a single instruction concurrently on multiple data.}
      \label{F:SIMD}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMT.pdf}
      \caption{Schematic of SIMT processing. A single compute unit (e.g., a GPU streaming multiprocessor) contains many processing elements and hosts many threads, each with an instruction to execute (I1, I2).  Threads with the same instruction execute concurrently on multiple data while the others must wait (leading to thread divergence).}
      \label{F:SIMT}
  \end{subfigure}
\end{figure}

@@

SIMT processing---the foundation of modern GPUs~\cite{lindholm2008nvidia}---is a related computing paradigm that hosts large groups of threads on a single compute element (a streaming multiprocessor in NVIDIA terminology).
A group of threads---typically \num{32} threads, known as a warp on NVIDIA GPUs---execute the same SIMD instruction on multiple data concurrently (\cref{F:SIMT}).
If some threads must execute a different instruction---e.g., due to if\slash then branching or predication---they are forced to wait and execute later.
This phenomena, known as thread-divergence, is a key consideration for SIMT-processing and can cause serious performance degradation for complicated algorithms~\cite{CurtisGPU:2017}.

@@

Yonkee and Sutherland~\cite{Yonkee2016} implemented SIMT-accelerated evaluations of thermodynamic parameters, multicomponent transport properties and species production rates for solution of a partial differential equations (PDEs) on an Intel Xeon E5-2680 CPU and an Tesla K20 GPU.

@@

This work will build upon our previous analytical chemical kinetic Jacobian code, \texttt{pyJac}~\cite{Niemeyer:2016aa}, in order to:
\begin{itemize}
 \item Derive and validate a new Jacobian formulation that greatly increases sparsity
 \item Enable cross-platform SIMD\slash SIMT vectorization for the CPU, GPU and other accelerators
 \item Investigate the performance for a wide range chemical kinetic models, specifically looking at speedups due to SIMD\slash SIMT-vectorization as well as relative to the previous version of \texttt{pyJac}, and finally
 \item Discuss future extensions to this work as well as the most promising directions for SIMD\slash SIMT vectorization in reactive-flow simulations
\end{itemize}
.

>>>>

As the importance of detailed chemical kinetics for predictive reactive-flow simulations has become recognized~\cref{LU2009192}, the size and complexity of chemical kinetic models describing current and next-generation fuels relevant to transportation and power generation have increased greatly; e.g., a recent biodiesel model consists of \textasciitilde\num{3500} chemical species, and over \num{17000} reactions.
Moreover, the cost of evaluating the chemical source terms scales linearly with the size of the model, while evaluation and factorization of the chemical kinetic Jacobian, if naively implemented, scale quadratically and cubically with the number of species in the model~\cref{LU2009192}.
These factors often make the use of detailed chemical kinetics prohibitive in practice; for direct numerical simulations, evaluation of reaction rates can consume around half of the total runtime~\cite{Spafford:2010aa}.
In addition, chemical kinetic Jacobian evaluation and factorization---required for most implicit integration techniques utilized to deal with numerical stiffness---becomes a bottleneck for even modestly sized chemical models in realistic reactive flow simulations~\cite{LU2009192}, and necessitates use of other cost-reduction strategies.

Single-Instruction, Multiple-Data (SIMD) and the related Single-Instruction, Multiple-Thread (SIMT) processing are two important vector-processing paradigms used increasingly in scientific computing.
Traditional multi-core parallelism is often used to increase central processing unit (CPU) performance, however as exponential growth in processing power---colloquially known as Moore's law---has slowed~\cite{khan2018science}, SIMD processors as well as SIMT processors, e.g., in the form of graphics processing units (GPUs), have gained recognition due to their increased floating operation throughput.
The parallel programming standard OpenCL~\cite{stone2010opencl} has further enabled adoption of vector-processing based codes in scientific computing by providing a common application program interface (API) for execution on heterogeneous systems, e.g., the CPU, GPU, or Intel's Many Integrated Core architecture (MIC), etc.
In this discussion, we will largely use OpenCL terminology to describe these processing paradigms, as it provides a convenient way to classify otherwise disparate processor type (e.g., CPUs and GPUs), however the concepts discussed are broadly applicable to SIMD\slash SIMT processing and not tied to OpenCL specifically.

\begin{figure}[phtb]
  \centering
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMD.pdf}
      \caption{Schematic of SIMD processing.  A single compute unit (e.g., a CPU core) contains a vector unit with $N_v$ processing elements, together called a vector-lane.  The vector unit executes a single instruction concurrently on multiple data.}
      \label{F:SIMD}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\linewidth}
      \includegraphics[width=\textwidth]{SIMT.pdf}
      \caption{Schematic of SIMT processing. A single compute unit (e.g., a GPU streaming multiprocessor) contains many processing elements and hosts many threads, each with an instruction to execute (I1, I2).  Threads with the same instruction execute concurrently on multiple data while the others must wait (leading to thread divergence).}
      \label{F:SIMT}
  \end{subfigure}
\end{figure}

@@

SIMT processing---the foundation of modern GPUs~\cite{lindholm2008nvidia}---is a related computing paradigm that hosts large groups of threads on a single compute element (a streaming multiprocessor in NVIDIA terminology).
A group of threads---typically \num{32}, known as a warp on NVIDIA GPUs---execute the same SIMD instruction on multiple data concurrently (\cref{F:SIMT}).
If some threads must execute a different instruction---e.g., due to if\slash then branching or predication---they are forced to wait and execute later.
This phenomena, known as thread-divergence, is a key consideration for SIMT-processing and can cause serious performance degradation for complicated algorithms~\cite{CurtisGPU:2017}.

@@

Yonkee and Sutherland~\cite{Yonkee2016} implemented SIMT-accelerated evaluations of thermodynamic parameters, multicomponent transport properties and species production rates for solution of a partial differential equations (PDEs) on an Intel Xeon E5-2680 CPU and a Tesla K20 GPU.

@@

This work will build upon our previous analytical chemical kinetic Jacobian code, \texttt{pyJac}~\cite{Niemeyer:2016aa}, in order to:
\begin{itemize}
 \item Derive and validate a new Jacobian formulation that greatly increases sparsity
 \item Enable cross-platform SIMD\slash SIMT vectorization for the CPU, GPU and other accelerators
 \item Investigate the performance for a wide range chemical kinetic models, specifically looking at speedups due to SIMD\slash SIMT-vectorization as well as relative to the previous version of \texttt{pyJac}, and finally
 \item Discuss future extensions to this work as well as the most promising directions for SIMD\slash SIMT vectorization in reactive-flow simulations
\end{itemize}
